---
title: "RMD Midterm"
author: "Bonnie N. Young et al."
date: "10/07/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load necessary packages
# install.packages("rnoaa")
# install.packages("riem")
# install.packages("usmap")
# install.packages("rgeos")
# install.packages("sp")
# install.packages("rgdal")
# install.packages("Hmisc")
# install.packages("broom")
# install.packages("dlnm")

# Run libraries
library(tidyverse)
library(data.table)
library(lubridate)
library(dplyr)
library(rnoaa)
library(riem)
library(usmap)
library(rgeos)
library(sp)
library(rgdal)
library(Hmisc)
library(splines)
library(broom)
library(dlnm)

# Load data: merged and cleaned (previously by GK)
# getwd()
setwd("C:/Users/bonyoung/Desktop/Advanced Epi Class - Temp/Midterm Report/R Midterm Project/data")

# Get data and filter out any obs with missing death data
data <- readRDS("timeseries_df.rds") %>%
  filter(!is.na(deaths))

```

## Old code to merge 3 datasets
> Old code to pull and merge Covid, EPA, and NOAA datasets.

```{r, "Old code data merge"}
####Read in COVID-19 data####
#covid_df <- read.csv("us-counties.csv") %>%
#  filter(county=="Los Angeles",
#         state=="California") %>%
#  mutate(date=as.Date(date,format="%Y-%m-%d")) %>%
#  arrange(date)

####Read in cleaned NOAA data####
#weather_df <- readRDS("weather_df_summ.rds") %>%
#  filter(county=="LA") %>%
#  mutate(county=ifelse(county=="LA","Los Angeles",NA))

####Read in cleaned exposure data####
#exp_df <- readRDS("total_exp_data.rds") %>%
#  filter(county=="Los Angeles")

####Merge together all three dataframes####
#timeseries_df <- full_join(covid_df,weather_df,
#                           by=c("county","date")) %>%
 # full_join(exp_df,by=c("county","date"))
#saveRDS(timeseries_df,"timeseries_df.rds")
```


## Get daily death counts
```{r}

# Loop to calculate daily deaths as a count per day from cumulative deaths
for(i in 1:nrow(data)) {
  if (i==1) {
    data[i,"daily_deaths"] <- 0
  } else {
    data[i,"daily_deaths"] <- data[i,"deaths"]-data[i-1,"deaths"]
  }
}

View(data%>%select(date, deaths, daily_deaths))

# Total deaths 26,013 equal new count of deaths per day
sum(data$daily_deaths)==max(data$deaths)
```

## Filter data set for cleaning
```{r}

# Create new clean dataset after removing death outlier of n=930 from feb 24, 2021; remove n=8 days of negative death counts (i.e., reporting errors); remove missing exposure and temperature variables
ts_mod_df <- data %>%
  filter(daily_deaths>-1) %>%
  filter(daily_deaths<900) %>%
  filter(!is.na(PM_conc_mean),
         !is.na(Ozone_conc_mean),
         !is.na(max_temp))

```

## Mutate date to weekday
```{r}

# Converting date to day of week
ts_mod_df <- ts_mod_df %>%
  mutate(dow=weekdays(date))

# Identify dow as a factor
ts_mod_df$dow <- factor(ts_mod_df$dow,ordered=FALSE)

```


## Mutate ozone to ug/m3 * 2140
```{r}

# Converting ozone ppm to ug/m3
ts_mod_df <- ts_mod_df %>%
  mutate(Ozone_conc_mean=Ozone_conc_mean*2140)

```

## Data summary
```{r, "Data summary"}

head(ts_mod_df)  
summary(ts_mod_df$PM_conc_mean)
summary(ts_mod_df$Ozone_conc_mean, na.rm=TRUE)
summary(ts_mod_df$max_temp, na.rm=TRUE)

# Calc standard deviations
sd(ts_mod_df$PM_conc_mean, na.rm=TRUE)
sd(ts_mod_df$Ozone_conc_mean, na.rm=TRUE)
sd(ts_mod_df$max_temp, na.rm=TRUE)

# Pearson correlation coefficients
ts_mod_df %>% 
  select(PM_conc_mean, Ozone_conc_mean, max_temp) %>% 
  cor(use = "complete.obs") 
```

## Plot exposures and outcome
``` {r, "Plot exp and outcome"}

# Look for long-term trends in the exposure? In the outcome?  
# Plot trends in deaths, PM, and ozone over study period with a  smooth line with geom_smooth to help detec longer-term patterns

# Covid deaths cumulative totals
ggplot(ts_mod_df, aes(x = date, y = deaths)) + 
  geom_point(alpha = 0.5, size = 0.5) + 
  labs(x = "Date", 
       y = "COVID-19 cumulative deaths") + 
  theme_bw()

# Covid deaths daily counts
ggplot(ts_mod_df, aes(x = date, y = daily_deaths)) + 
  geom_line(alpha = 0.5, size = 0.5) + 
  geom_smooth()+
  labs(x = "Date", 
       y = "COVID-19 daily deaths") + 
  theme_bw()

# PM
ggplot(ts_mod_df, aes(x = date, y = PM_conc_mean)) + 
  geom_point(alpha = 0.5, size = 0.5) + 
  geom_smooth()+
  labs(x = "Date", 
       y = "PM2.5 concentration") + 
  theme_bw()

# Ozone
ggplot(ts_mod_df, aes(x = date, y = Ozone_conc_mean)) + 
  geom_point(alpha = 0.5, size = 0.5) + 
  geom_smooth()+
  labs(x = "Date", 
       y = "Ozone concentration") + 
  theme_bw()

# Temperature
ggplot(ts_mod_df, aes(x = date, y = max_temp)) + 
  geom_point(alpha = 0.5, size = 0.5) + 
  geom_smooth()+
  labs(x = "Date", 
       y = "Temperature") + 
  theme_bw()

```

## GLM  modeling analysis
> Code help from Grace Kuiper

### Linear model with daily PM2.5 predicting daily deaths
```{r}
#' linear model with daily PM2.5 predicting daily deaths
mod_linear_reg <- glm(daily_deaths ~ PM_conc_mean, data=ts_mod_df,
                      family="quasipoisson")

summary(mod_linear_reg)

glance(mod_linear_reg)
tidy(mod_linear_reg,conf.int=T)
augment(mod_linear_reg)

mod_linear_reg %>% 
  augment() %>%
  ggplot(aes(x=PM_conc_mean)) +
  geom_point(aes(y=daily_deaths),alpha=0.4,size=0.5) +
  geom_line(aes(y=.fitted),color="red") +
  labs(x="Mean daily PM2.5", y="Expected mortality count") +
  theme_bw()
```

### Linear model with daily ozone predicting daily deaths
```{r}
#' linear model with daily PM2.5 predicting daily deaths
mod_linear_reg_o <- glm(daily_deaths ~ Ozone_conc_mean, data=ts_mod_df,
                      family="quasipoisson")

summary(mod_linear_reg_o)

glance(mod_linear_reg_o)
tidy(mod_linear_reg_o,conf.int=T)
augment(mod_linear_reg_o)

mod_linear_reg_o %>% 
  augment() %>%
  ggplot(aes(x=Ozone_conc_mean)) +
  geom_point(aes(y=daily_deaths),alpha=0.4,size=0.5) +
  geom_line(aes(y=.fitted),color="red") +
  labs(x="Mean daily ozone", y="Expected mortality count") +
  theme_bw()

```

### Linear model with daily max temperature predicting daily deaths
```{r}

#' linear model with temp predicting daily deaths
mod_linear_reg_t <- glm(daily_deaths ~ max_temp, data=ts_mod_df,
                      family="quasipoisson")

summary(mod_linear_reg_t)

glance(mod_linear_reg_t)
tidy(mod_linear_reg_t,conf.int=T)
augment(mod_linear_reg_t)

mod_linear_reg_t %>% 
  augment() %>%
  ggplot(aes(x=max_temp)) +
  geom_point(aes(y=daily_deaths),alpha=0.4,size=0.5) +
  geom_line(aes(y=.fitted),color="red") +
  labs(x="Max temp (F)", y="Expected mortality count") +
  theme_bw()

```

### Linear model with daily Pm2.5 AND ozone predicting daily deaths
```{r}
mod_linear_reg_2 <- glm(daily_deaths ~ PM_conc_mean + Ozone_conc_mean, data=ts_mod_df,
                      family="quasipoisson")

summary(mod_linear_reg_2)

glance(mod_linear_reg_2)
tidy(mod_linear_reg_2,conf.int=T)
augment(mod_linear_reg_2)

mod_linear_reg_2 %>% 
  augment() %>%
  ggplot(aes(x=PM_conc_mean)) +
  geom_point(aes(y=daily_deaths),alpha=0.4,size=0.5) +
  geom_line(aes(y=exp(.fitted)),color="red") +
  labs(x="Mean daily PM2.5", y="Log(Expected mortality count)") +
  theme_bw()

mod_linear_reg_2 %>%
  augment() %>%
  ggplot(aes(x=Ozone_conc_mean)) +
  geom_point(aes(y=daily_deaths),alpha=0.4,size=0.5) +
  geom_line(aes(y=exp(.fitted)),color="red") +
  labs(x="Daily maximum 8-hour ozone",y="Log(Expected mortality count)") +
  theme_bw()
```


###' linear model with daily PM2.5, control for ozone, dow, temp
```{r}
mod_linear_reg_3 <- glm(daily_deaths ~ PM_conc_mean + Ozone_conc_mean + dow + max_temp, data=ts_mod_df,
                        family="quasipoisson")

summary(mod_linear_reg_3)

glance(mod_linear_reg_3)
tidy(mod_linear_reg_3,conf.int=T)
augment(mod_linear_reg_3)

mod_linear_reg_3 %>% 
  augment() %>%
  ggplot(aes(x=PM_conc_mean)) +
  geom_point(aes(y=daily_deaths),alpha=0.4,size=0.5) +
  geom_line(aes(y=exp(.fitted)),color="red") +
  labs(x="Mean daily PM2.5", y="Log(Expected mortality count)") +
  theme_bw()

```

###' How are daily mortality related with a linear and non-linear function of time?
```{r}
ts_mod_df <- ts_mod_df %>%
  mutate(time=as.numeric(date)-first(as.numeric(date)))

ggplot(ts_mod_df, aes(x=time,y=daily_deaths)) +
  geom_point(size=0.5,alpha=0.5) +
  theme_bw()

mod_time <- glm(daily_deaths ~ time, data=ts_mod_df, family="quasipoisson")
mod_time %>%
  augment() %>%
  ggplot(aes(x=time)) +
  geom_point(aes(y=daily_deaths), alpha=0.4,size=0.5) +
  geom_line(aes(y=exp(.fitted)),color="red") +
  labs(x="Date", y="Expected mortality count") +
  theme_bw()

# library(splines)
mod_time_nonlin <- glm(daily_deaths ~ ns(time,df=12),
                       data=ts_mod_df,family="quasipoisson")
mod_time_nonlin %>%
  augment() %>%
  mutate(time=ts_mod_df$time) %>%
  ggplot(aes(x=time)) +
  geom_point(aes(y=daily_deaths),alpha=0.4,size=0.5) +
  geom_line(aes(y=exp(.fitted)),color="red") +
  labs(x="Date",y="Expected mortality count") +
  theme_bw()
```

###'Now fit model with ozone and pm, controlling for DOW and long-term and seasonal trends in mortality using spline with 12 df on time
```{r}

# Controlling for only dow
mod_ctrl_dow <- glm(daily_deaths ~ PM_conc_mean + Ozone_conc_mean + dow, 
                    data = ts_mod_df, family = "quasipoisson")
#mod_ctrl_dow %>% 
#  tidy() %>% 
#  filter(grepl("conc_mean",term))


# Controlling for dow and time as a spline df=12
mod_ctrl_dow_time <- glm(daily_deaths ~ PM_conc_mean + Ozone_conc_mean + dow +
                           ns(time, df = 12), 
                         data = ts_mod_df, family = "quasipoisson")

mod_ctrl_dow_time %>%
  tidy() %>%
  filter(grepl("conc_mean",term))

tidy(mod_ctrl_dow_time,conf.int=T)
```

###'Add in controlling for temperature
```{r}
mod_ctrl_weather <- glm(daily_deaths ~ PM_conc_mean + Ozone_conc_mean + dow +
                           max_temp + ns(time, df = 12), 
                         data = ts_mod_df, family = "quasipoisson")

summary(mod_ctrl_weather)

mod_ctrl_weather %>% 
  tidy() %>% 
  filter(grepl("conc_mean",term))

tidy(mod_ctrl_weather,conf.int=T)

mod_ctrl_weather %>% 
  augment() %>%
  ggplot(aes(x=PM_conc_mean)) +
  geom_point(aes(y=daily_deaths),alpha=0.4,size=0.5) +
  geom_line(aes(y=exp(.fitted)),color="red") +
  labs(x="Mean daily PM2.5", y="Log(Expected mortality count)") +
  theme_bw()
```

###'Include non-linear functions on PM, ozone, and temperature
```{r}
mod_ctrl_nl <- glm(daily_deaths ~ ns(PM_conc_mean,df=4) + ns(Ozone_conc_mean,df=4) + dow +
                          ns(max_temp,df=4) + ns(time, df = 12), 
                        data = ts_mod_df, family = "quasipoisson")

summary(mod_ctrl_nl)
tidy(mod_ctrl_nl,conf.int=T)

mod_ctrl_nl %>%
  augment() %>%
  mutate(PM_conc_mean = ts_mod_df$PM_conc_mean) %>%
  ggplot(aes(x=PM_conc_mean)) +
  geom_point(aes(y=daily_deaths),alpha=0.4,size=0.5) +
  geom_line(aes(y=exp(.fitted)),color="red",alpha=0.7,size=0.4) +
  labs(x="Daily mean PM2.5",y="Expected mortality count") +
  theme_bw()

mod_ctrl_nl %>% 
  tidy() %>% 
  filter(grepl("conc_mean",term))
```

###'Generate predictions using spline for PM2.5 and center the reference value for RR at WHO annual mean guideline of 5 ug/m3.
```{r}
pm_spline <- ns(ts_mod_df$PM_conc_mean, df=4)
pm_spline_preds <- predict(pm_spline, newx=2.5:63.2)
pm_spline_preds %>%
  head()

mod_ctrl_nl %>%
  tidy() %>%
  filter(str_detect(term,"PM_conc_mean"))

pm_spline_ests <- mod_ctrl_nl %>%
  tidy() %>%
  filter(str_detect(term,"PM_conc_mean")) %>%
  pull(estimate)

pm_spline_ests

pred_pm_function <- tibble(
  pm = 2.5:63.2,
  pm_func=pm_spline_preds %*% pm_spline_ests,
  rr=exp(pm_func)
)

pred_pm_function %>%
  head()

ggplot(pred_pm_function,aes(x=pm,y=rr)) +
  geom_point() +
  geom_line() +
  labs(x="PM2.5",
       y="Relative Risk") +
  geom_hline(yintercept = 1.0,color="red",linetype=2) +
  geom_vline(xintercept = 5.0,color="blue",linetype=3) +
  geom_vline(xintercept = 15.0,color="blue",linetype=3) +
  theme_bw()

# library(dlnm)
pm_basis <- crossbasis(ts_mod_df$PM_conc_mean, lag = 0, 
                         argvar = list(fun = "ns", df = 4), 
                         arglag = list(fun = "integer"))
pm_basis %>% 
  head()
pm_spline %>%
  head()

dlnm_mod_1 <- glm(daily_deaths ~ pm_basis + dow + ns(max_temp,df=4) +
                    ns(Ozone_conc_mean,df=4) + ns(time,df=12),
                  data=ts_mod_df,family="quasipoisson")
dlnm_mod_1 %>%
  tidy() %>%
  filter(str_detect(term,"pm_basis"))

# Center the reference of RR to 5 ug/m3
crosspred(basis=pm_basis,model=dlnm_mod_1,
          cen=5,by=1) %>%
  str()

est_rr <- dlnm_mod_1 %>%
  crosspred(basis=pm_basis,model=.,cen=5,by=1) %>%
  pluck("allRRfit")
est_rr

dlnm_temp_function <- tibble(PM_conc_mean=as.numeric(names(est_rr)),
                             rr=est_rr)

dlnm_temp_function %>%
  head()

# Show full set of values for RR for entire range of PM
dlnm_temp_function

ggplot(dlnm_temp_function,aes(x=PM_conc_mean,y=rr)) +
  geom_point() +
  geom_line() +
  labs(x="Mean daily PM2.5",y="Relative Risk") +
  geom_hline(yintercept = 1.0,color="red",linetype=2)+
  theme_bw() +
  geom_vline(xintercept = 5.0,color="blue",linetype=3)+
  geom_vline(xintercept=15.0,color="blue",linetype=3)
dlnm_mod_1 %>%
  crosspred(basis=pm_basis,model=.,cen=5,by=1) %>%
  plot()
```





# SENSITIVITY ANALYSES

### Reduce data set for sens analysis
> Filter to only data from June 1,2020 - Dec 31, 2020, n=213 obs

```{r}
 
ts_mod_df_filtered <- ts_mod_df %>% filter(date>"2020-05-31",date<"2021-01-01")


```


### With reduced dataset, include non-linear functions on PM, ozone, and temperature
> Reduce degrees of freedom because only 7 months of data now.

```{r}

# Reduce df for time ns df=4
mod_ctrl_nl <- glm(daily_deaths ~ ns(PM_conc_mean,df=4) + ns(Ozone_conc_mean,df=4) + dow +
                          ns(max_temp,df=4) + ns(time, df = 4), 
                        data = ts_mod_df_filtered, family = "quasipoisson")

summary(mod_ctrl_nl)
tidy(mod_ctrl_nl,conf.int=T)

mod_ctrl_nl %>%
  augment() %>%
  mutate(PM_conc_mean = ts_mod_df_filtered$PM_conc_mean) %>%
  ggplot(aes(x=PM_conc_mean)) +
  geom_point(aes(y=daily_deaths),alpha=0.4,size=0.5) +
  geom_line(aes(y=exp(.fitted)),color="red",alpha=0.7,size=0.4) +
  labs(x="Daily mean PM2.5",y="Expected mortality count") +
  theme_bw()

mod_ctrl_nl %>% 
  tidy() %>% 
  filter(grepl("conc_mean",term))
```

## USE reduced dataset and same  code to generate predictions using spline for PM2.5 and center the reference value for RR at WHO annual mean guideline of 5 ug/m3.
```{r}

pm_spline <- ns(ts_mod_df_filtered$PM_conc_mean, df=4)
pm_spline_preds <- predict(pm_spline, newx=2.5:63.2)
pm_spline_preds %>%
  head()

mod_ctrl_nl %>%
  tidy() %>%
  filter(str_detect(term,"PM_conc_mean"))

pm_spline_ests <- mod_ctrl_nl %>%
  tidy() %>%
  filter(str_detect(term,"PM_conc_mean")) %>%
  pull(estimate)

pm_spline_ests

pred_pm_function <- tibble(
  pm = 2.5:63.2,
  pm_func=pm_spline_preds %*% pm_spline_ests,
  rr=exp(pm_func)
)

pred_pm_function %>%
  head()

ggplot(pred_pm_function,aes(x=pm,y=rr)) +
  geom_point() +
  geom_line() +
  labs(x="PM2.5",
       y="Relative Risk") +
  geom_hline(yintercept = 1.0,color="red",linetype=2) +
  geom_vline(xintercept = 5.0,color="blue",linetype=3) +
  geom_vline(xintercept = 15.0,color="blue",linetype=3) +
  theme_bw()

# library(dlnm)
pm_basis <- crossbasis(ts_mod_df_filtered$PM_conc_mean, lag = 0, 
                         argvar = list(fun = "ns", df = 4), 
                         arglag = list(fun = "integer"))
pm_basis %>% 
  head()
pm_spline %>%
  head()

dlnm_mod_1 <- glm(daily_deaths ~ pm_basis + dow + ns(max_temp,df=4) +
                    ns(Ozone_conc_mean,df=4) + ns(time,df=4),
                  data=ts_mod_df_filtered,family="quasipoisson")
dlnm_mod_1 %>%
  tidy() %>%
  filter(str_detect(term,"pm_basis"))

# Center reference value for RR at 5 ug/m3:
crosspred(basis=pm_basis,model=dlnm_mod_1,
          cen=5,by=1) %>%
  str()

est_rr <- dlnm_mod_1 %>%
  crosspred(basis=pm_basis,model=.,cen=5,by=1) %>%
  pluck("allRRfit")
est_rr

dlnm_temp_function <- tibble(PM_conc_mean=as.numeric(names(est_rr)),
                             rr=est_rr)

dlnm_temp_function %>%
  head()

# Show full set of values for RR for entire range of PM
dlnm_temp_function

ggplot(dlnm_temp_function,aes(x=PM_conc_mean,y=rr)) +
  geom_point() +
  geom_line() +
  labs(x="Mean daily PM2.5",y="Relative Risk") +
  geom_hline(yintercept = 1.0,color="red",linetype=2)+
  theme_bw() +
  geom_vline(xintercept = 5.0,color="blue",linetype=3)+
  geom_vline(xintercept=15.0,color="blue",linetype=3)
dlnm_mod_1 %>%
  crosspred(basis=pm_basis,model=.,cen=5,by=1) %>%
  plot()

```


